---
title: "hw3"
author: "Senna"
date: "2024-10-12"
output: github_document
---
```{r}
library(tidyverse)
library(lubridate)
```

### Problem 1
```{r}
library(p8105.datasets)
p1_df = ny_noaa
```

Do some data cleaning. Create separate variables for year, month, and day.  
Ensure observations for temperature, precipitation, and snowfall are given in reasonable units. For snowfall, what are the most commonly observed values? Why?
```{r}
p1_df = p1_df|>
  mutate(
    year = year(date),
    month = month(date),
    day = day(date)
  )

p1_df <- p1_df|>
  mutate(
    tmax = as.integer(tmax),
    tmin = as.integer(tmin)
  )

```

```{r}
################# idk how to do this wth
summary(p1_df[["tmax"]])
summary(p1_df[["tmin"]]) #-40c to 50c
summary(p1_df[["prcp"]]) # in mm or inches (0-few hundred mm)
summary(p1_df[["snow"]]) # in cm or inches
```
```{r}
snow_mode <- p1_df |>
  group_by(snow) |>
  summarize(count = n()) |>
  arrange(desc(count))


head(snow_mode,3)
```



Make a two-panel plot showing the average max temperature in January and in July in each station across years. Is there any observable / interpretable structure? Any outliers?

Make a two-panel plot showing (i) tmax vs tmin for the full dataset (note that a scatterplot may not be the best option); and (ii) make a plot showing the distribution of snowfall values greater than 0 and less than 100 separately by year.



### problem 2


Load, tidy, merge, and otherwise organize the data sets. Your final dataset should include all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes (i.e. not numeric, and using factors with the ordering of tables and plots in mind).

```{r}
accel_df = read_csv(file = "./nhanes_accel.csv")|>
  janitor::clean_names()

demo_df = read_csv(file = "./nhanes_covar.csv", skip=4)|>
  janitor::clean_names()|>
  filter (age >= 21)|>
  drop_na()|>
  mutate(
    education = factor(education),
    sex = recode(sex,
                 '1'= 'M',
                 '2' = 'F'))

#sex: binary??????????????? 


p2_df = full_join(demo_df, accel_df, by='seqn')
```


Produce a reader-friendly table for the number of men and women in each education category, 
```{r}
education_table = p2_df|>
  group_by (education, sex)|>
  summarise (count = n(), .groups = 'drop')|>
  pivot_wider (
    names_from = sex,
    values_from = count
  )|>
  mutate(
    education = recode(
      education,
      '1' = 'less than high school',
      '2' = 'high school equivalent',
      '3' = 'more than high school'
    )
  )|>
  arrange(education)

education_table
```

and create a visualization of the age distributions for men and women in each education category. Comment on these items.
```{r}
age_distribution_plot <- ggplot(p2_df, aes(x = education, y = age, fill = sex)) +
  geom_boxplot(position = position_dodge(0.8)) +
  labs(title = "Age Distributions by Education Category and Sex",
       x = "Education Category",
       y = "Age",
       fill = "Sex") +
  theme_minimal()

# Display the plot
print(age_distribution_plot)
```



total activity over the day : aggregate across minutes to create a total activity variable for each participant. Plot these total activities (y-axis) against age (x-axis); your plot should compare men to women and have separate panels for each education level. Include a trend line or a smooth to illustrate differences. Comment on your plot.
```{r}

min1_index = which (colnames(p2_df)== 'min1')
last_column_index <- ncol(p2_df)


total_activity <- p2_df |>
  mutate(total_activity = rowSums(across(min1_index:last_column_index), na.rm = TRUE)) |>
  select(seqn, sex, age, bmi, education, total_activity)

```
```{r}
total_activity|>
  ggplot(aes( x = age, y = total_activity, color = as.factor(sex)))+
  geom_point()+
  geom_smooth(se= FALSE)+
  facet_wrap( ~ education, labeller = as_labeller(c(
    '1' = 'Less than high school',
    '2' = 'High school equivalent',
    '3' = 'More than high school'
  ))) +
  labs(
    title = "Total Activity vs. Age by Education Level",
    x = "Age",
    y = "Total Activity",
    color = "Sex"
  ) +
  theme_minimal() +
  theme(legend.position = "right")
```


inspection of activity over the course of the day. Make a three-panel plot that shows the 24-hour activity time courses for each education level and use color to indicate sex. Describe in words any patterns or conclusions you can make based on this graph; including smooth trends may help identify differences.

```{r}
activity_long_df <- p2_df|>
  pivot_longer(
    cols = starts_with("min"),
    names_to = "minute",
    values_to = "activity"
  ) |>
  mutate(
    minute = as.numeric(gsub("min", "", minute))  
  )


ggplot(activity_long_df, aes(x = minute, y = activity, color = as.factor(sex))) +
  geom_smooth( se = FALSE) + # Smooth trend line
  facet_wrap(~ education, labeller = as_labeller(c(
    '1' = 'Less than high school',
    '2' = 'High school equivalent',
    '3' = 'More than high school'
  ))) +
  labs(
    title = "24-Hour Activity Time Courses by Education Level and Sex",
    x = "Minute of the Day",
    y = "Activity Level",
    color = "Sex"
  ) +
  scale_x_continuous(breaks = seq(0, 1440, by = 120), labels = function(x) paste0(x / 60, "h")) + 
  theme_minimal() +
  theme(legend.position = "top")
```



### Problem 3

















